---
title: "Hypothesis test and P-value"
description: |
  In this post we are going to test our hypothesis by observing the differences between different stats and the impact of a variable on another. 
author:
  - name: Guilherme Bastos Gomes
    url: https://gbastosg.github.io/guilhermesportfolio/
date: 2022-05-22
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, layout="body-outset", preview=TRUE}
library(tidyverse)

ToothGrowth$dose <- as.factor(ToothGrowth$dose)

dose_05 <- ToothGrowth %>%
              filter(dose == 0.5)

dose_1 <- ToothGrowth %>%
              filter(dose == 1)

observable_diff1 <- 19.735 - 10.605
observable_diff2 <- 26.100	- 19.735
observable_diff3 <- 26.100	- 10.605

n <- 10000
null <- vector("numeric",n)
for (i in 1:n) {
media_controle <- ToothGrowth$len %>% sample(12) %>% mean()
media_tratamento <- ToothGrowth$len %>% sample(12) %>% mean()
null[i] <- mean(media_tratamento) - mean(media_controle)
}

hist(null, freq=TRUE)

abline(v=observable_diff1, col="red", lwd=2)

abline(v=observable_diff2, col="blue", lwd=2)

```

# Hello!

In this post I am explaining how to perform a hypothesis test using R, and we also going to talk about what is a p-value that is always appearing on scientifical writing.

# ToothGrowth Data set

In this in-built data set (remember to call `tidiverse`) called `ToothGrowth` we can find a total of 60 observations divided into 3 variables:

- len
- supp
- dose

```{r, layout="l-body-outset"}
library(rmarkdown)
paged_table(ToothGrowth)
```

This data was obtained via research, I've explained this data with more details in [this post](), but generally it demonstrates the effect of applied vitamin C in odontoblasts growing (cells responsible for teeth growing) in guinea pigs, with different doses.

```
ggplot(ToothGrowth) +
  geom_freqpoly(aes(x = len, color = dose), binwidth = 1) +
  theme_classic()
```
ggplot(ToothGrowth) +
  geom_freqpoly(aes(x = len, color = dose), binwidth = 1) +
  theme_classic()


```{r, layout="l-body-outset", preview=TRUE}
library(ggthemes)
ggplot(ToothGrowth) +
  geom_freqpoly(aes(x = len, color = dose), binwidth = 1) +
  theme_economist()
```

It seems like there is a clear influence of the vitamin dose applied in the guinea pigs and their odontoblasts sizes, but we need to be skepticals and obtain some evidences over what are we claiming before taking this hypothesis under consideration.

# At first we look at the means

To start understanding the relation of the vitamin doses with tooth growth, we can observe and compare the means of the groups using the `summarise()` function of the `dplyr` package.

```
ToothGrowth %>%
  group_by(dose) %>%
  summarise(mean = mean(len))
```
```{r, layout="body-outset", preview=TRUE}
library(tidyverse)
ToothGrowth %>%
  group_by(dose) %>%
  summarise(mean = mean(len))
```

It seems like there is also a difference between the mean of each dose of vitamin applied, let's also analyse the difference between means:

```
observable_diff1 <- 19.735 - 10.605
observable_diff2 <- 26.100	- 19.735
observable_diff3 <- 26.100	- 10.605
```

Until here we have already answered some questions:

- Would the teeth of rats that received a bigger dose of vitamin C, frequently larger?

```
#Another form of obtaining variation per dose

dose_05 <- ToothGrowth %>%
              filter(dose == 0.5)

dose_1 <- ToothGrowth %>%
              filter(dose == 1)
              
dose_2 <- ToothGrowth %>%
              filter(dose == 2)
```

We can check if this is `TRUE` with the following code:

```
#Inside of observable_diff1 we have the difference between the 0.5 and 1 doses

mean(dose_1$len) - mean(dose_05$len) == observable_diff1
```

```{r, layout="l-body-outset", preview=TRUE}
mean(dose_1$len) - mean(dose_05$len) == observable_diff1
```

- Which are the largest odontoblasts of each groups? And the smallest?

```
ToothGrowth %>%
  group_by(dose) %>%
  summarise(min = min(len), max = max(len))
```

```{r, layout="l-body-outset", preview=TRUE}
ToothGrowth %>%
  group_by(dose) %>%
  summarise(min = min(len), max = max(len))
```

So this is it, it really seems like the dose of Vitamin C has the direct effect over teeth growth, well actually not... what really happens is that if we repeat the experiment with different rat groups, we probably would have obtained a different answer from the one we obtained previously.

FOr this reason, this variables are known by `random variables`, they can take several different values.

That means, everytime we repeat the experiment, we will have different means to the groups we have access to (dose), and by being so the variable `len` is considered `random`.

Let's have a better understandment of `random variables`.

# Random Variables

Let's comprehend how the means change when we get other samples of our data set:

```
control <- ToothGrowth$len %>% sample(12)
```

```{r, layout="l-body-outset", preview=TRUE}
control <- ToothGrowth$len %>% sample(12)
```

We can do this repeatedly to build a completely random distribution based on our main data (mainly because we don't have the means to repeat the experiment, so we model what would be the population without doses of vitamin C being applied to them)

Please notice how the mean changes in each sample, repeat the command of sampling in your R using the function `sample()`.

```
mean_control <- ToothGrowth$len %>% sample(12) %>% mean()

print(media_controle)

mean_control <- ToothGrowth$len %>% sample(12) %>% mean()

print(mean_control)
```

```{r, layout="l-body-outset", preview=TRUE}
mean_control <- ToothGrowth$len %>% sample(12) %>% mean()

print(media_controle)

mean_control <- ToothGrowth$len %>% sample(12) %>% mean()

print(mean_control)
```

Notice how the mean varies in each measument, we can do this many times to learn even more about the distribution of this random variable.

As previously said we have to be skepticals, for that we will assume that randomly the observed difference could not mean much, that is it could be something that happened by chance.

To obtain evidences and minimise this uncertainty, we will use the `Null hypothesis`.

# Null hypothesis

Let's analyse again the observed differences of before. As we need to be skepticals we have to question ourselves:

- How can we affirm that the observed differences are happenning because of the applied doses?
- What would happen if we gave all the rats the same dose?
- Would we obsserve differences as large as these ones?

In this moment we are working with the `Null hypothesis`, which name comes to remind us that we are acting as skepticals: we give credit to the possibility that our observations could just have happened by chance.

Null, because we are questioning what would happen to the observable difference in case there were no application in the experiment, that is in case nothing have happened, what would happen?

## Building a control group to test the Null hypothesis

We can check, how many times we like, the differences between the sizes of the teeth without comparing doses. We can do that by sampling a random number of rats from our original data set, and then we can record the value of the differences between means of two random groups, the script for that I learned with prof. Rafa, in the book "Data analysis for the life sciences":

```
mean_control <- ToothGrowth$len %>% sample(12) %>% mean()

mean_treatment <- ToothGrowth$len %>% sample(12) %>% mean()

print(mean_treatment - mean_control)
```

It's also possible to do this may times using loops in R:

```
n <- 10000 
null <- vector("numeric",n) #first we create a vector with 10000 spaces
for (i in 1:n) { #looping to generate a null vector
mean_control <- ToothGrowth$len %>% sample(12) %>% mean()
mean_treatment <- ToothGrowth$len %>% sample(12) %>% mean()
null[i] <- mean(mean_treatment) - mean(mean_control)
}

print(null)
```

```{r, layout="l-body-outset", preview=TRUE}
n <- 10000 
null <- vector("numeric",n)
for (i in 1:n) {
mean_control <- ToothGrowth$len %>% sample(12) %>% mean()
mean_treatment <- ToothGrowth$len %>% sample(12) %>% mean()
null[i] <- mean(mean_treatment) - mean(mean_control)
}
```

The values in  `null` are what we call a null distribution.

What percentage of  `null` is equal or higher than the observed differences?

```
mean(null >= observable_diff1)

mean(null >= observable_diff2)

mean(null >= observable_diff3)
```

```{r, layout="body-outset", preview=TRUE}
mean(null >= observable_diff1)

mean(null >= observable_diff2)

mean(null >= observable_diff3)
```

### Interpreting results

Parece que apenas uma pequena porcentagem dentre as nossas 10.000 simulações é igual ou maior as diferenças observadas na hipótese nula. Como céticos o que concluimos? Quando não existe efeito de doses, podemos observar uma diferença tão grande quanto observamos antes apenas 0,0001% das vezes (no máximo com a `diferenca_observavel1`). Esse valor é conhecido como p-value ou p-valor.

# P-value ou P-valor

Fizemos uma simulação para contruir o data set `null`, vamos observar as diferenças observaveis caso não houvessem doses aplicadas:

```
hist(null, freq=TRUE)
```

```{r, layout="l-body-outset", preview=TRUE}
hist(null, freq=TRUE)
```

Podemos observar também quais valores são tão maiores quanto as nossas diferenças observadas:


```
hist(null, freq=TRUE)

abline(v=diferenca_observavel1, col="red", lwd=2)

abline(v=diferenca_observavel2, col="blue", lwd=2)

abline(v=diferenca_observavel3, col="yellow", lwd=2)
```

```{r, layout="l-body-outset", preview=TRUE}
hist(null, freq=TRUE)

abline(v=diferenca_observavel1, col="red", lwd=2)

abline(v=diferenca_observavel2, col="blue", lwd=2)

abline(v=diferenca_observavel3, col="yellow", lwd=2)
```

Os valores maiores ou iguais as diferenças observáveis são relativamente raros. O que nos garante maior evidência a respeito da nossa hipótese de que a quantidade de vitamina C afeta o crescimento dos odontoblastos.

Foi possível perceber que a maior diferença, entre as doses 0.5 e 2, nem aparecem em nossa distribuição nula, mostrando que essa diferença observável não ocorreu ao acaso e que a hipótese nula pode ser refutada.

Mas ainda assim devemos tomar cuidado!

# Grupo controle a partir de outras doses

Para criar um grupo controle com apenas uma dose, podemos filtrar o data set original

```
ToothGrowth_dot5 <- ToothGrowth %>%
  filter(dose == 0.5)
```

```{r, layout="l-body-outset", preview=TRUE}
ToothGrowth_dot5 <- ToothGrowth %>%
  filter(dose == 0.5)
```

Agora podemos criar nossos data sets para construir a hipótese nula:

```
controle_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()

tratamento_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()

print(tratamento_ToothGrowth_dot5 - controle_ToothGrowth_dot5)
```

```{r, layout="l-body-outset", preview=TRUE}
controle_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()

tratamento_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()

print(tratamento_ToothGrowth_dot5 - controle_ToothGrowth_dot5)
```

Aqui repetimos 10.000 vezes a diferença entre as médias para criar uma distribuição nula:

```
n <- 10000 
null_dose <- vector("numeric",n) #primeiro criamos um vetor com 10000 espaços
for (i in 1:n) { #looping para criar um vetor nulo
controle_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()
tratamento_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()
null_dose[i] <- mean(tratamento_ToothGrowth_dot5) - mean(controle_ToothGrowth_dot5)
}

print(null)
```
```{r, layout="l-body-outset", preview=TRUE}
n <- 10000 
null_dose <- vector("numeric",n) #primeiro criamos um vetor com 10000 espaços
for (i in 1:n) { #looping para criar um vetor nulo
controle_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()
tratamento_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()
null_dose[i] <- mean(tratamento_ToothGrowth_dot5) - mean(controle_ToothGrowth_dot5)
}
```

Analisando a diferença observada em relação a esse data set, usando a cor amarela para marcar a primeira diferença observável, entre as médias das dose 0.5 e 1, e vermelho para usar a segunda, entre as médias de 1 e 2:

```
hist(null_dose, freq=TRUE)

```

```{r, layout="l-body-outset", preview=TRUE}
hist(null_dose, freq=TRUE)

```

# Usando o pacote infer()

O pacote infer nos ajuda a testar hipóteses de uma forma bem fácil e direta
Instale e carregue o pacote:

```
install.packages("infer")

library(infer)
```

Para calcular a estatística da diferença observada:

```
F_hat <- ToothGrowth %>% 
  specify(len ~ dose) %>%
  calculate(stat = "F")
```
```{r, layout="l-body-outset", preview=TRUE}
library(infer)
F_hat <- ToothGrowth %>% 
  specify(len ~ dose) %>%
  calculate(stat = "F")
```
Gerando a distribuição nula:

```
null_dist <- ToothGrowth %>% 
   specify(len ~ dose) %>%
   hypothesize(null = "independence") %>%
   generate(reps = 1000, type = "permute") %>%
   calculate(stat = "F")
```
```{r, layout="l-body-outset", preview=TRUE}
null_dist <- ToothGrowth %>% 
   specify(len ~ dose) %>%
   hypothesize(null = "independence") %>%
   generate(reps = 1000, type = "permute") %>%
   calculate(stat = "F")
```
Visualizando a diferença observada com a distribuição nula:

```
visualize(null_dist) +
  shade_p_value(obs_stat = F_hat, direction = "greater")
```
```{r, layout="l-body-outset", preview=TRUE}
visualize(null_dist) +
  shade_p_value(obs_stat = F_hat, direction = "greater")
```

Falaremos ainda mais sobre isso em outros posts quando usaremos todo esse conhecimento aplicado a um projeto! Até mais!
