---
title: "Hypothesis test and P-value"
description: |
  In this post we are going to test our hypothesis by observing the differences between different stats and the impact of a variable on another. 
author:
  - name: Guilherme Bastos Gomes
    url: https://gbastosg.github.io/guilhermesportfolio/
date: 2022-05-22
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, layout="body-outset", preview=TRUE}
library(tidyverse)

ToothGrowth$dose <- as.factor(ToothGrowth$dose)

dose_05 <- ToothGrowth %>%
              filter(dose == 0.5)

dose_1 <- ToothGrowth %>%
              filter(dose == 1)

observable_diff1 <- 19.735 - 10.605
observable_diff2 <- 26.100	- 19.735
observable_diff3 <- 26.100	- 10.605

n <- 10000
null <- vector("numeric",n)
for (i in 1:n) {
media_controle <- ToothGrowth$len %>% sample(12) %>% mean()
media_tratamento <- ToothGrowth$len %>% sample(12) %>% mean()
null[i] <- mean(media_tratamento) - mean(media_controle)
}

hist(null, freq=TRUE)

abline(v=observable_diff1, col="red", lwd=2)

abline(v=observable_diff2, col="blue", lwd=2)

```

# Hello!

In this post I am explaining how to perform a hypothesis test using R, and we also going to talk about what is a p-value that is always appearing on scientifical writing.

# ToothGrowth Data set

In this in-built data set (remember to call `tidiverse`) called `ToothGrowth` we can find a total of 60 observations divided into 3 variables:

- len
- supp
- dose

```{r, layout="l-body-outset"}
library(rmarkdown)
paged_table(ToothGrowth)
```

This data was obtained via research, I've explained this data with more details in [this post](), but generally it demonstrates the effect of applied vitamin C in odontoblasts growing (cells responsible for teeth growing) in guinea pigs, with different doses.

```
ggplot(ToothGrowth) +
  geom_freqpoly(aes(x = len, color = dose), binwidth = 1) +
  theme_classic()
```
ggplot(ToothGrowth) +
  geom_freqpoly(aes(x = len, color = dose), binwidth = 1) +
  theme_classic()


```{r, layout="l-body-outset", preview=TRUE}
library(ggthemes)
ggplot(ToothGrowth) +
  geom_freqpoly(aes(x = len, color = dose), binwidth = 1) +
  theme_economist()
```

It seems like there is a clear influence of the vitamin dose applied in the guinea pigs and their odontoblasts sizes, but we need to be skepticals and obtain some evidences over what are we claiming before taking this hypothesis under consideration.

# At first we look at the means

To start understanding the relation of the vitamin doses with tooth growth, we can observe and compare the means of the groups using the `summarise()` function of the `dplyr` package.

```
ToothGrowth %>%
  group_by(dose) %>%
  summarise(mean = mean(len))
```
```{r, layout="body-outset", preview=TRUE}
library(tidyverse)
ToothGrowth %>%
  group_by(dose) %>%
  summarise(mean = mean(len))
```

It seems like there is also a difference between the mean of each dose of vitamin applied, let's also analyse the difference between means:

```
observable_diff1 <- 19.735 - 10.605
observable_diff2 <- 26.100	- 19.735
observable_diff3 <- 26.100	- 10.605
```

Until here we have already answered some questions:

- Would the teeth of rats that received a bigger dose of vitamin C, frequently larger?

```
#Another form of obtaining variation per dose

dose_05 <- ToothGrowth %>%
              filter(dose == 0.5)

dose_1 <- ToothGrowth %>%
              filter(dose == 1)
              
dose_2 <- ToothGrowth %>%
              filter(dose == 2)
```

We can check if this is `TRUE` with the following code:

```
#Inside of observable_diff1 we have the difference between the 0.5 and 1 doses

mean(dose_1$len) - mean(dose_05$len) == observable_diff1
```

```{r, layout="l-body-outset", preview=TRUE}
mean(dose_1$len) - mean(dose_05$len) == observable_diff1
```

- Which are the largest odontoblasts of each groups? And the smallest?

```
ToothGrowth %>%
  group_by(dose) %>%
  summarise(min = min(len), max = max(len))
```

```{r, layout="l-body-outset", preview=TRUE}
ToothGrowth %>%
  group_by(dose) %>%
  summarise(min = min(len), max = max(len))
```

So this is it, it really seems like the dose of Vitamin C has the direct effect over teeth growth, well actually not... what really happens is that if we repeat the experiment with different rat groups, we probably would have obtained a different answer from the one we obtained previously.

FOr this reason, this variables are known by `random variables`, they can take several different values.

That means, every time we repeat the experiment, we will have different means to the groups we have access to (dose), and by being so the variable `len` is considered `random`.

Let's have a better comprehension about `random variables`.

# Random Variables

Let's uderstand how the means change when we get samples of our data set:

```
control <- ToothGrowth$len %>% sample(12)
```

```{r, layout="l-body-outset", preview=TRUE}
control <- ToothGrowth$len %>% sample(12)
```

We can do this repeatedly to build a completely random distribution based on our main data (mainly because we don't have the means to repeat the real experiment, so we model what would be the population without doses of vitamin C being applied to them).

Please notice how the `mean` change in each sample, repeat the command of sampling in your R using `sample()`.

```
mean_control <- ToothGrowth$len %>% sample(12) %>% mean()

print(media_controle)

mean_control <- ToothGrowth$len %>% sample(12) %>% mean()

print(mean_control)
```

```{r, layout="l-body-outset", preview=TRUE}
mean_control <- ToothGrowth$len %>% sample(12) %>% mean()

print(media_controle)

mean_control <- ToothGrowth$len %>% sample(12) %>% mean()

print(mean_control)
```

Notice how the mean varies in each measurement, we can do this many times to learn even more about the distribution of this random variable.

As previously said we have to be skeptical, for that we will assume that randomly the observed difference could not mean much, that is it could be something that happened by chance.

To obtain evidences and minimise this uncertainty, we will use the `Null hypothesis`.

# Null hypothesis

Let's analyse again the observed differences of before. As we need to be skepticals we have to question ourselves:

- How can we affirm that the observed differences are happenning because of the applied doses?
- What would happen if we gave all the rats the same dose?
- Would we observe differences as large as these ones?

In this moment we are working with the `Null hypothesis`, which name comes to remind us that we are acting as skeptical: we give credit to the possibility that our observations could just have happened by chance.

Null, because we are questioning what would happen to the observable difference in case there were no application in the experiment, that is in case nothing have happened, what would happen?

## Building a control group to test the Null hypothesis

We can check, how many times we like, the differences between the sizes of the teeth without comparing doses. We can do that by sampling a random number of rats from our original data set, and then we can record the value of the differences between means of two random groups, the script for that I learned with prof. Rafa, in the book "Data analysis for the life sciences":

```
mean_control <- ToothGrowth$len %>% sample(12) %>% mean()

mean_treatment <- ToothGrowth$len %>% sample(12) %>% mean()

print(mean_treatment - mean_control)
```

It's also possible to do this may times using loops in R:

```
n <- 10000 
null <- vector("numeric",n) #first we create a vector with 10000 spaces
for (i in 1:n) { #looping to generate a null vector
mean_control <- ToothGrowth$len %>% sample(12) %>% mean()
mean_treatment <- ToothGrowth$len %>% sample(12) %>% mean()
null[i] <- mean(mean_treatment) - mean(mean_control)
}

print(null)
```

```{r, layout="l-body-outset", preview=TRUE}
n <- 10000 
null <- vector("numeric",n)
for (i in 1:n) {
mean_control <- ToothGrowth$len %>% sample(12) %>% mean()
mean_treatment <- ToothGrowth$len %>% sample(12) %>% mean()
null[i] <- mean(mean_treatment) - mean(mean_control)
}
```

The values in  `null` are what we call a null distribution.

What percentage of  `null` is equal or higher than the observed differences?

```
mean(null >= observable_diff1)

mean(null >= observable_diff2)

mean(null >= observable_diff3)
```

```{r, layout="body-outset", preview=TRUE}
mean(null >= observable_diff1)

mean(null >= observable_diff2)

mean(null >= observable_diff3)
```

### Interpreting results

Only a minority percentage within our 10.000 simulations are equal or higher than the observed differences in the null hypothesis. As skepticals what do we conclude? That when there's no dose effect, we can observe a difference as big as the one observed before only 0,0001% of the times (in maximum, as could see in `observable_diff1`). This value is known as p-value.

# P-value

We performed a simulation to build the `null` data set, let's now observe the the null distribution:

```
hist(null, freq=TRUE)
```

```{r, layout="l-body-outset", preview=TRUE}
hist(null, freq=TRUE)
```

We can also see where are the values as high as our observations:

```
hist(null, freq=TRUE)

abline(v=observable_diff1, col="red", lwd=2)

abline(v=observable_diff2, col="blue", lwd=2)

abline(v=observable_diff3, col="yellow", lwd=2)
```

```{r, layout="l-body-outset", preview=TRUE}
hist(null, freq=TRUE)

abline(v=observable_diff1, col="red", lwd=2)

abline(v=observable_diff2, col="blue", lwd=2)

abline(v=observable_diff3, col="yellow", lwd=2)
```

Values higher than those ones we have observed are relatively rare. This gives us evidence to support our hypothesis, that the amount of dose of vitamin c applied in guinea pigs affects the size of odontoblasts.

It was possible to notice that the top difference, between the doses 0.5 and 2,
 didn't even appeared in our plot of the null distribution, showing us that the observable difference couldn't have happened entirely randomly, and that the null hypothesis can be overthrown.
 
But even though, we should always be carefull!

# Creating a control group using a filter

To create a control group with only one dose, we can filter the data set:

```
ToothGrowth_dot5 <- ToothGrowth %>%
  filter(dose == 0.5)
```

```{r, layout="l-body-outset", preview=TRUE}
ToothGrowth_dot5 <- ToothGrowth %>%
  filter(dose == 0.5)
```

Now we can create our data sets to build the null hypothesis:

```
controle_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()

tratamento_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()

print(tratamento_ToothGrowth_dot5 - controle_ToothGrowth_dot5)
```

```{r, layout="l-body-outset", preview=TRUE}
controle_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()

tratamento_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()

print(tratamento_ToothGrowth_dot5 - controle_ToothGrowth_dot5)
```

Here we have repeated 10.000 times the differences between means to create a null distribution:

```
n <- 10000 
null_dose <- vector("numeric",n)
for (i in 1:n) { 
controle_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()
tratamento_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()
null_dose[i] <- mean(tratamento_ToothGrowth_dot5) - mean(controle_ToothGrowth_dot5)
}

print(null)
```

```{r, layout="l-body-outset", preview=TRUE}
n <- 10000 
null_dose <- vector("numeric",n) #primeiro criamos um vetor com 10000 espaços
for (i in 1:n) { #looping para criar um vetor nulo
controle_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()
tratamento_ToothGrowth_dot5 <- ToothGrowth_dot5$len %>% sample(12) %>% mean()
null_dose[i] <- mean(tratamento_ToothGrowth_dot5) - mean(controle_ToothGrowth_dot5)
}
```

```
hist(null_dose, freq=TRUE)

```

```{r, layout="l-body-outset", preview=TRUE}
hist(null_dose, freq=TRUE)

```

# Using the infer() package

The `infer` package helps us to test hypothesis in a pretty straight-forward way, install the package and load it:

```
install.packages("infer")

library(infer)
```

To calculate the observed statistic in our data set:

```
F_hat <- ToothGrowth %>% 
  specify(len ~ dose) %>%
  calculate(stat = "F")
```
```{r, layout="l-body-outset", preview=TRUE}
library(infer)
F_hat <- ToothGrowth %>% 
  specify(len ~ dose) %>%
  calculate(stat = "F")
```

Generating a null distribution:

```
null_dist <- ToothGrowth %>% 
   specify(len ~ dose) %>%
   hypothesize(null = "independence") %>%
   generate(reps = 1000, type = "permute") %>%
   calculate(stat = "F")
```
```{r, layout="l-body-outset", preview=TRUE}
null_dist <- ToothGrowth %>% 
   specify(len ~ dose) %>%
   hypothesize(null = "independence") %>%
   generate(reps = 1000, type = "permute") %>%
   calculate(stat = "F")
```

Visualising the observable difference with the calculated null distribution:

```
visualize(null_dist) +
  shade_p_value(obs_stat = F_hat, direction = "greater")
```
```{r, layout="l-body-outset", preview=TRUE}
visualize(null_dist) +
  shade_p_value(obs_stat = F_hat, direction = "greater")
```

We will talk even more about this in other posts, when I will use all of this knowledge applied in a project! See you!
