---
title: "About the structure of variables"
description: |
  Variables are everywhere, let's understand a bit more about their structures in the R programming language.
author:
  - name: Guilherme Bastos Gomes
    url: https://gbastosg.github.io/guilhermesportfolio/
date: 2022-04-19
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, layout="body-outset", preview=TRUE}
library(tidyverse)
library(ggthemes)

mpg %>%
  ggplot(aes(x = displ, y = cty, color = as.factor(displ))) +
  geom_point() +
  facet_grid(vars(year)) +
  theme_bw() +
  labs(
    title = "Motor efficiency from different car displays",
    subtitle = "by: Guilherme Bastos",
    caption = "Source: mpg data set",
    x = "Display",
    y = "City miles per gallon") +
  theme(
        legend.title = element_blank(),
        )
  
  

```


# Hello! In this post I am going to briefly write about the structure of variables

>A variable is a quantity, quality or property that you can measure. 

But what is it in practice?

Everytime we create an object, we are also creating a kind of variable

In fact, it's good to remember that:

> $\color{blue}{\text{"Everything that exists in R is an object."}}$
~John M. Chambers

So basically we will be dealing with variables all the time!

To understand the 4 main types of variables, you should create the following objects in your  workspace:

* logical <- `TRUE`
* character <- `"text"`
* numeric <- `1`
* factor <- as.factor(c("I", "Am", "a", "factor"))

Each object is interpreted in a different way by the language, let's use the function `class()` to understand how each variable behaves:

# class()

Type in the following after creating the objects:

```
class(logical)
class(character)
class(numeric)
class(factor)
```
You should have the following answers:

```
[1] "logical"
[1] "character"
[1] "numeric"
[1] "factor"
```

Each object belongs to a different class, therefore should be treated as such.

# What each of these structures means?

## logical

* **logical (ou lógica)**: Boolean/binary , it's always `TRUE` or `FALSE`

## character

* **character (or string)**: simple text, including symbols and numbers inside of quotation marks `"text"`, `"this is a text"`, `"th1s 1s 4ls0 a t3xt"`.

## numeric

* **numeric (numérico)**: Literally numbers. Mathematic operators can be used here.

Remembering that in truth programming languages are just *fancy calculators*:

* `+` add
* `-` subtract
* `*` multiply
* `/` divide
* `^` exponentiation

## factor

* **factor**: Numbers or strings (text) ORDERED somehow:

This is a factor:
```
as.factor(c("I", "am", "a", "factor"))
```
This is also a factor:
```
as.factor(c(1, 2, 3, 4))
```

The word ORDER is the most important one here to define factors

These are the main 4 structures that will appear in your analyses!

#### Thank you for your time!
Follow me on twitter: @gimbgomes


